Headgroup structure and cation binding in phosphatidylserine lipid bilayers

### 2019-05-27 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

Experimental studies have indicated that the
PS headgroup is more rigid than the phosphatidylcholine (PC)
owing to electrostatic interactions or the formation of hydrogen bonding network between the headgroups~\cite{browning80,buldt81}.
While most monovalent ions interact weakly with
PS-containing bilayers, multivalent cations and Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
The dehydrated complexes of PS headgroups with calcium ions can even lead to
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}. Mixing PS lipids with PC lipids reduces their propensity to form strong complexes with multivalent ions and makes the PS headgroup less rigid~\cite{browning80,buldt81,roux90,roux91}.
That being said, some studies suggest Ca$^{2+}$ has similar specific binding affinity 
to negatively charged and zwiterionic phospholipids, and that
the increased cation binding to negatively charged lipid bilayers arises only due
to the increased local cation concentration in the vicinity of the membrane~\cite{seelig90,sinn06}.

@@

The molecular level interpretation of these observations is currently lacking, and classical molecular dynamics (MD) simulations have been widely used in efforts 
to understand the PS headgroup structure, its influence on lipid bilayer properties, and its
interaction with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
Unfortunately, the results have depended on the force field used.
For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}. In contrast,
the results from the same lipid model without the NBfix ion parameters, combined with NMR chemical shifts and
rotational-echo double-resonance (REDOR) experiments, indicate a significant binding affinity also to the phosphate region \cite{hallock18}.
Meanwhile, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
combined with fluorescent and vibrational sum frequency spectroscopy, suggest substantial
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.

We have recently demonstrated that the lipid C--H bond order parameters, $S_\mathrm{CH}$,
can be utilized to resolve such controversies~\cite{botan15,catte16}. The $S_\mathrm{CH}$ can be
measured from NMR experiments with high accuracy and directly compared to simulations
in order to evaluate the quality of the simulation model, or to interpret the experiments \cite{ollila16}. Using this approach,
it has been established that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most MD force fields~\cite{botan15}, that the cation binding to PC
lipid bilayers is overestimated~\cite{catte16}, and that the splitting of the order parameters at
carbon 2 in {\it sn}-2 lipid tails is correctly reproduced only in the CHARMM36 force field~\cite{piggot17}.% [https://pubs.acs.org/doi/abs/10.1021/acs.jctc.7b00643].
%
%Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
%\todo{Should we leave the mention of ECC out from the Introduction
%(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H\,NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences substantial forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available. 
%Moved here from the molecular electrometer paragraph because there it distrupted the flow of the section and it contains results. This also enables referring to figures in order. I also removed the refs which do not seem to refer to the actual data used in the figure.
%Rewrote the paragraph, see if ok.

@@

The different PS MD models produce highly varied  headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the SI ),
as previously observed also for PC lipids \cite{botan15}. Moreover, none of the models produces a set of order parameters in full agreement with the experiments.
The models perform generally less well for PS than for PC
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\cite{botan15}).
This complicates the interpretation of structural differences between PC and PS headgroups. However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36-UA) do replicate the larger-than-in-PC
forking of the $\alpha$-carbon observed in the experiments. Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}), in line with
experiments (Fig. \ref{HGorderParameters}).

@@

These results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}, and the suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments. However, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to confirm the correct conformational ensemble.

@@

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers compared to PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups. We explore this by quantifying the counterion density profiles along the membrane normal accompanied by the areas per lipid in Fig.~\ref{NAdensPOPS}.
The force fields studied show significant differences in both binding affinity
and distribution of ions at the interface.
The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
is reproduced only in GROMOS-CKP and in the MacRog simulation
with potassium counterions, while other models give considerably smaller values (Fig.~\ref{NAdensPOPS}).
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
since the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36-UA simulation, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
The results are in line with the previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.

@@

Experimentally, the headgroup order parameters of POPC
increase when negatively charged POPS lipids are incorporated in lipid bilayer
(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
The CHARMM36 and Berger simulations either exhibit no change or show a decrease
in the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
Therein, the stronger counterion binding cancels
the effect of negatively charged headgroups and prevents the experimentally observed
increase of headgroup order parameters with growing amount of PS lipids.
Therefore, we suggest that the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces the most
realistic surface charge density in membranes containing PS lipids,
while the other tested models overestimate the counterion
binding affinity. The results are in line with the behaviour of headgroup order
parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
in the SI.

@@

Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
exhibit a strong dependence of CaCl$_2$ at small concentrations yet they saturate already at a concentration below 100~mM (Fig. \ref{changesWITHCaClPS}).
The $\beta$-carbon order parameter increases with added CaCl$_2$,
whereas the larger $\alpha$-carbon order parameter decreases. Moreover, a slight increase is observed in
the smaller $\alpha$-carbon value. All these changes are significantly overestimated in the
tested simulation models---including CHARMM36---which underestimated the binding affinity.
Importantly, the different simulation models predict qualitatively different behaviour
for the two POPS $\alpha$-carbon order parameters as a function of added calcium.
For example, both order parameters decrease in Berger, but increase
in MacRog, whereas Lipid14/17 and CHARMM36 models exhibit more complex behavior.
This is in contrast to the PC headgroup, where a
qualitatively correct response to bound ions is reproduced
by all simulation models, despite the significant discrepancies in the headgroup
structures observed in salt-free simulations~\cite{catte16}.
The divergent response of the Berger model may arise from the ring like structures
observed in the headgroup region in this model (Fig. 6 in Ref. \citenum{mukhopadhyay04}).
%These ring like structures are a widespread feature of typical Berger based lipid force
%fields containing explicit hydrogen atoms in the head group
%[REFS https://pubs.acs.org/doi/abs/10.1021/jp900645z and https://www.sciencedirect.com/science/article/pii/S0300908408000692 and https://pubs.acs.org/doi/abs/10.1021/jp9110019]
Therefore, we conclude that
the improvement of force fields is necessary to correctly capture the interactions between the
PS headgroup and calcium ions in MD simulations.

@@

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
% Put your acknowledgments here.
  MJ acknowledges financial support from the Emil Aaltonen foundation and
  and CSC-IT center for science for computational resources.
  OHSO acknowledges financial support from Academy of Finland (315596),
  Integrated Structural Biology Research Infrastructure of
  Helsinki Institute of Life Science (Instruct-HiLIFE), and
  CSC-IT center for science for computational resources.
  TJP wishes to acknowledge the use of the Iridis computing resources at the University of Southampton.
  FFR acknowledges Tecnol\'{o}gico Nacional de M\'{e}xico Campus Zacatecas Occidente and
  Direcci\'{o}n General de Asuntos del Personal Acad\'{e}mico (DGAPA)
  Programa de Apoyo a Proyectos de Investigaci\'{o}n e Innovaci\'{o}n
  Tecnol\'{o}gica (PAPIIT) IG100416 for financial support and Cl\'{u}ster
  H\'{i}brido de Superc\'{o}mputo Xiuhcoatl-CINVESTAV and Miztli-Direcci\'{o}n de
  C\'{o}mputo y de Tecnolog\'{i}as de Informaci\'{o}n y Comunicaci\'{o}n (DGTIC)-
  Universidad Nacional Aut\'{o}noma de M\'{e}xico (UNAM) (Project
  LANCAD-UNAM-DGTIC-028) facilities for computing-time allocation.
\end{acknowledgments}

>>>>

Experimental studies have indicated that the
PS headgroup is more rigid than the phosphatidylcholine (PC)
owing to electrostatic interactions or the formation of a hydrogen bonding network between the headgroups~\cite{browning80,buldt81}.
While most monovalent ions interact weakly with
PS-containing bilayers, multivalent cations and Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
The dehydrated complexes of PS headgroups with calcium ions can even lead to
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}. Mixing PS lipids with PC lipids reduces their propensity to form strong complexes with multivalent ions and makes the PS headgroup less rigid~\cite{browning80,buldt81,roux90,roux91}.
That being said, some studies suggest Ca$^{2+}$ has similar specific binding affinity 
to negatively charged and zwiterionic phospholipids, and that
the increased cation binding to negatively charged lipid bilayers arises only due
to the increased local cation concentration in the vicinity of the membrane~\cite{seelig90,sinn06}.

@@

The molecular level interpretation of these observations is currently lacking, and classical molecular dynamics (MD) simulations have been widely used in efforts 
to understand the PS headgroup structure, its influence on lipid bilayer properties, and its
interaction with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
Unfortunately, the results have depended on the particular force field used.
For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
the CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}. In contrast,
the results from the same lipid model without the NBfix ion parameters, combined with NMR chemical shifts and
rotational-echo double-resonance (REDOR) experiments, indicate a significant binding affinity also toward the phosphate region \cite{hallock18}.
Meanwhile, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
combined with fluorescent and vibrational sum frequency spectroscopy, suggest substantial
calcium binding to the carbonyls in the acyl chains \cite{melcrova16}.

We have recently demonstrated that the lipid C--H bond order parameters, $S_\mathrm{CH}$,
can be utilized to resolve such controversies~\cite{botan15,catte16}. The $S_\mathrm{CH}$ can be
measured from NMR experiments with high accuracy and directly compared to simulations
in order to evaluate the quality of the simulation model and/or to interpret the experiments \cite{ollila16}. Using this approach,
it has been established that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most MD force fields~\cite{botan15}, that the cation binding to PC
lipid bilayers is overestimated~\cite{catte16}, and that the splitting of the order parameters at
carbon 2 in {\it sn}-2 lipid tails is correctly reproduced only in the CHARMM36 force field~\cite{piggot17}.% [https://pubs.acs.org/doi/abs/10.1021/acs.jctc.7b00643].
%
%Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
%\todo{Should we leave the mention of ECC out from the Introduction
%(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H\,NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences substantial forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but a detailed structrural interpretation is not
available. 
%Moved here from the molecular electrometer paragraph because there it distrupted the flow of the section and it contains results. This also enables referring to figures in order. I also removed the refs which do not seem to refer to the actual data used in the figure.
%Rewrote the paragraph, see if ok.

@@

The different PS MD models produce highly varied headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the SI ),
as previously observed also for PC lipids \cite{botan15}. Moreover, none of the models produces a set of order parameters in full quantitative agreement with the experiments.
Generally, the models perform less well for PS than for PC
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\cite{botan15}).
This complicates the interpretation of structural differences between the PC and PS headgroups. However, concentrating on the headgroup alone, we see that the best performing models (Slipids, CHARMM36 and CHARMM36-UA) indeed do replicate the larger-than-in-PC
forking of the $\alpha$-carbon that is observed in the experiments. Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}), in line with
experiments (Fig. \ref{HGorderParameters}).

@@

These results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}, and the suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments. However, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to elucidate the correct conformational ensemble.

@@

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers compared to PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups. We explore this by quantifying the counterion density profiles along the membrane normal accompanied by the areas per lipid in Fig.~\ref{NAdensPOPS}.
The force fields studied show significant differences in both binding affinity
and distribution of ions at the interface.
The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
is reproduced only in GROMOS-CKP and in the MacRog simulation
with potassium counterions, while other models give considerably smaller values (Fig.~\ref{NAdensPOPS}).
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
since the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36-UA simulation, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
The results are consistent with the previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.
%Would be potentially interesting to flesh out what the role of the hydrating water molecules play in the specific lipid headgroup/ion configurations that are sampled in simulations. My feeling is that "fully hydrated" lipids might carry with them differing amounts of water between the different FFs.. -JJM

@@

Experimentally, the headgroup order parameters of POPC
increase when negatively charged POPS lipids are incorporated in lipid bilayer
(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
The CHARMM36 and Berger simulations either exhibit no change or show a decrease
in the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
Therein, the stronger counterion binding cancels
the effect of negatively charged headgroups and prevents the experimentally observed
increase of headgroup order parameters with growing amount of PS lipids.
Therefore, we suggest that the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces the most
realistic surface charge density in membranes containing PS lipids,
while the other tested models overestimate the counterion
binding affinity. The results are consistent with the behavior of headgroup order
parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
in the SI.

@@

Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
exhibit a strong dependence of CaCl$_2$ at small concentrations yet they saturate already at a concentration below 100~mM (Fig. \ref{changesWITHCaClPS}).
The $\beta$-carbon order parameter increases with added CaCl$_2$,
whereas the larger $\alpha$-carbon order parameter decreases. Moreover, a slight increase is observed in
the smaller $\alpha$-carbon value. All these changes are significantly overestimated in the
tested simulation models---including CHARMM36---which underestimated the binding affinity.
Importantly, the different simulation models predict qualitatively different behavior
for the two POPS $\alpha$-carbon order parameters as a function of added calcium.
For example, both order parameters decrease in Berger, but increase
in MacRog, whereas Lipid14/17 and CHARMM36 models exhibit more complex behavior.
This is in contrast to the PC headgroup, where a
qualitatively correct response to bound ions is reproduced
by all simulation models, despite the significant discrepancies in the headgroup
structures observed in salt-free simulations~\cite{catte16}.
The divergent response of the Berger model may arise from the ring like structures
observed in the headgroup region in this model (Fig. 6 in Ref. \citenum{mukhopadhyay04}).
%These ring like structures are a widespread feature of typical Berger based lipid force
%fields containing explicit hydrogen atoms in the head group
%[REFS https://pubs.acs.org/doi/abs/10.1021/jp900645z and https://www.sciencedirect.com/science/article/pii/S0300908408000692 and https://pubs.acs.org/doi/abs/10.1021/jp9110019]
Therefore, we conclude that
the improvement of force fields is necessary to correctly capture the interactions between the
PS headgroup and calcium ions in MD simulations.

@@

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
% Put your acknowledgments here.
  MJ acknowledges financial support from the Emil Aaltonen foundation and
  and CSC-IT center for science for computational resources.
  JJM gratefully acknowledges financial support from the Carlsberg Foundation in the form of a postdoctoral fellowship while at the University of Chicago (grants CF15-0552, CF16-0639, and CF17-0783) and the research framework provided by the Research Computing Center at the University of Chicago.
  OHSO acknowledges financial support from Academy of Finland (315596),
  Integrated Structural Biology Research Infrastructure of
  Helsinki Institute of Life Science (Instruct-HiLIFE), and
  CSC-IT center for science for computational resources.
  TJP wishes to acknowledge the use of the Iridis computing resources at the University of Southampton.
  FFR acknowledges Tecnol\'{o}gico Nacional de M\'{e}xico Campus Zacatecas Occidente and
  Direcci\'{o}n General de Asuntos del Personal Acad\'{e}mico (DGAPA)
  Programa de Apoyo a Proyectos de Investigaci\'{o}n e Innovaci\'{o}n
  Tecnol\'{o}gica (PAPIIT) IG100416 for financial support and Cl\'{u}ster
  H\'{i}brido de Superc\'{o}mputo Xiuhcoatl-CINVESTAV and Miztli-Direcci\'{o}n de
  C\'{o}mputo y de Tecnolog\'{i}as de Informaci\'{o}n y Comunicaci\'{o}n (DGTIC)-
  Universidad Nacional Aut\'{o}noma de M\'{e}xico (UNAM) (Project
  LANCAD-UNAM-DGTIC-028) facilities for computing-time allocation.
\end{acknowledgments}
